---
title: "FA5_Group1_Quijano_DSC1105"
author: "Julian Philip S. Quijano"
date: "2026-04-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggplot2)
library(reshape2)
```

```{r}
data <- read.csv("adult.csv")
head(data)
```

```{r}
data <- data %>%
  mutate(education_group = case_when(
    education %in% c("Preschool", "1st-4th", "5th-6th") ~ "Elementary",
    education %in% c("7th-8th", "9th") ~ "Middle School",
    education %in% c("10th", "11th", "12th", "HS-grad") ~ "High School",
    education %in% c("Some-college", "Assoc-acdm", "Assoc-voc") ~ "College / Associate",
    education %in% c("Bachelors", "Masters", "Doctorate", "Prof-school") ~ "Advanced",
    TRUE ~ NA_character_
  ))

anyNA(data)

table_ed_inc <- table(data$education_group, data$income)

table_with_totals <- addmargins(table_ed_inc)

row_percent <- prop.table(table_ed_inc, margin = 1) * 100

combined <- cbind(table_ed_inc, round(row_percent, 2))

```
```{r}
apa_table <- as.data.frame.matrix(table_ed_inc) %>%
  mutate(Total = rowSums(.)) %>%
  mutate(`<=50K (%)` = paste0(`<=50K`, " (", round(`<=50K`/Total*100, 2), "%)"),
         `>50K (%)` = paste0(`>50K`, " (", round(`>50K`/Total*100, 2), "%)")) %>%
  select(`<=50K (%)`, `>50K (%)`, Total)

apa_table
```
```{r}
chisq_test = chisq.test(table(data$education_group, data$income))

chi_stat <- chisq_test$statistic   
chi_dof <- chisq_test$parameter   
chi_p <- chisq_test$p.value

print("A chi-square test of independence showed a significant association between education level and income, X^2(df = 4) = 3738.87, p < .001. Since p < .05, the relationship between education level and income is statistically significant.")
```

```{r}
tab_ed_inc <- table(data$education_group, data$income)
tab_ed_inc

chisq_test <- chisq.test(tab_ed_inc)

expected <- chisq_test$expected

std_res <- chisq_test$stdres

round(expected, 2)
round(std_res, 2)
```

```{r}
res_df <- melt(std_res)

ggplot(res_df, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 4) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Heatmap of Standardized Residuals",
    x = "Income",
    y = "Education Level",
    fill = "Std Residual"
  ) +
  theme_minimal()
```

#### a.
A chi-square test of independence revealed a significant association between education level and income, X^2(df = 4) = 3738.87, p < 0.05. This suggests that income distribution varies across education levels.


#### b.
All categories show significant deviations, but key ones:

Advanced (>50K): z = 59.04 → far more high-income individuals than expected

Advanced (<=50K): z = -59.04 → far fewer low-income individuals than expected

High School (<=50K): z = 34.67 → overrepresented in lower income

High School (>50K): z = -34.67 → underrepresented in higher income

Elementary & Middle School also strongly:

Overrepresented in <=50K

Underrepresented in >50K


#### c.
Higher education strongly predicts higher income — individuals in the Advanced category are dramatically overrepresented in the >50K group.

Lower education levels are concentrated in lower income brackets, especially Elementary and Middle School groups.

High School and College/Associate groups fall in between, but still skew toward <= 50K income.

